---
title: "EDA"
format: PDF
editor: visual
---

```{r}
library(tidyverse)
```

```{r}
penn_crash <- read.table("PennCrash.txt", header = TRUE, quote = "\"", sep = "")

penn_crash
```
```{r}
sum(penn_model_data$Treat == 1)
```
only 331 treated observations compared to 1,655 treated.
```{r}
# Visualize relationship for 2008 (pre-treatment state)
ggplot(penn_crash, aes(x = adt08, y = tot08, color = Treat)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Traffic Volume (2008) vs. Total Crashes (2008)",
       x = "Annual Average Daily Traffic (2008)", 
       y = "Total Crashes (2008)") +
  theme_minimal()
```
# Trend plot
```{r}
# Reshape the data to long format for time-series visualization
crash_long <- penn_crash %>%
  pivot_longer(cols = c("tot04", "tot08", "tot12"), 
               names_to = "year", 
               values_to = "crashes") %>%
  mutate(year = recode(year, "tot04" = 2004, "tot08" = 2008, "tot12" = 2012))

# Calculate mean crashes per year for each group
crash_summary <- crash_long %>%
  group_by(Treat, year) %>%
  summarise(mean_crashes = mean(crashes), .groups = 'drop')

# Plot the trend
ggplot(crash_summary, aes(x = factor(year), y = mean_crashes, color = Treat, group = Treat)) +
  geom_line(size = 1) +
  geom_point(size = 3) +
  labs(title = "Mean Crash Counts Over Time",
       x = "Year", y = "Average Crashes",
       color = "Group") +
  theme_minimal()
```
Prior to treatment (2004–2008), the trends are broadly similar, providing visual support for the parallel trends assumption. After 2008, treated sites exhibit a larger decline in crashes compared to untreated sites, suggesting a potential effect of rumble strip installation. However, this visual difference should be interpreted cautiously, as it does not account for differences in traffic exposure or roadway characteristics. We must consult regression results to see if the estimated treatment effect is not statistically significant.

# Crash Rates
```{r}
# 1. Calculate yearly rates
penn_crash <- penn_crash %>%
  mutate(rate04 = tot04 / (adt04 * length),
         rate08 = tot08 / (adt08 * length),
         rate12 = tot12 / (adt12 * length))

# 2. Summarize the mean rates by year and treatment
yearly_rate_summary <- penn_crash %>%
  group_by(Treat) %>%
  summarise(mean_rate04 = mean(rate04, na.rm = TRUE),
            mean_rate08 = mean(rate08, na.rm = TRUE),
            mean_rate12 = mean(rate12, na.rm = TRUE))

print(yearly_rate_summary)

# 3. Optional: Plot the trend of these rates
rate_long <- yearly_rate_summary %>%
  pivot_longer(cols = starts_with("mean_rate"), names_to = "year", values_to = "rate") %>%
  mutate(year = recode(year, "mean_rate04" = 2004, "mean_rate08" = 2008, "mean_rate12" = 2012))

ggplot(rate_long, aes(x = factor(year), y = rate, color = factor(Treat), group = Treat)) +
  geom_line(size = 1) +
  geom_point(size = 3) +
  labs(title = "Average Yearly Crash Rate",
       x = "Year", y = "Crashes per Traffic-Mile",
       color = "Treatment") +
  theme_minimal()
```
# Confounder Analsyis
```{r}
# Check for balance across key categorical variables
# We use 'pivot_longer' to create a faceted plot of all categorical variables
categorical_vars <- c("width", "speed", "shoulder")

penn_crash %>%
  select(Treat, all_of(categorical_vars)) %>%
  pivot_longer(cols = -Treat, names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = value, fill = factor(Treat))) +
  geom_bar(position = "fill") + # 'fill' makes them percentage bars for easier comparison
  facet_wrap(~variable, scales = "free") +
  labs(title = "Proportional Balance of Road Characteristics by Treatment",
       y = "Proportion of Sites", x = "Category", fill = "Treatment") +
  theme_minimal()
```
The proportional bar charts reveal that rumble strip installations were not assigned randomly with respect to site characteristics. Specifically, treatment sites appear disproportionately represented in certain categories—such as higher-speed zones and specific pavement widths—compared to control sites. Because these characteristics (speed, width, shoulder) are independently associated with crash risk, the lack of balance indicates that simple comparisons between the treatment and control groups may be biased by these pre-existing structural differences. Consequently, these variables must be included as covariates in the regression model to isolate the effect of the rumble strips from the influence of road geometry.

```{r}
# Visualize the distribution of crashes by road characteristics
penn_crash %>%
  select(tot12, width, speed, shoulder) %>%
  pivot_longer(cols = -tot12, names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = value, y = tot12)) +
  geom_boxplot(outlier.alpha = 0.3) +
  facet_wrap(~variable, scales = "free") +
  labs(title = "Distribution of 2012 Crashes by Road Feature",
       y = "Total Crashes (2012)", x = "Category") +
  theme_minimal()
```
The exploratory distribution of crash counts (Figure X) exhibited significant right-skewness and overdispersion, characterized by a large proportion of zeros and a long tail of high-crash outliers. This distribution necessitates the use of a Generalized Linear Model (specifically a Negative Binomial regression) rather than OLS, as the latter would fail to account for the discrete nature and overdispersion of the crash counts.

# parallel trends


