---
title: "modeling"
format: pdf
editor: visual
---

```{r}
library(MASS)
library(tidyverse)
library(tidyr)
library(dplyr)
```

```{r}
penn_crash <- read.table("PennCrash.txt", header = TRUE, quote = "\"", sep = "")

penn_crash
```

# Preparing Data
```{r}
# 1. Clean data: Factorize categorical predictors
penn_model_data <- penn_crash %>%
  mutate(
    Treat = factor(Treat),
    width = factor(width),
    speed = factor(speed),
    shoulder = factor(shoulder)
  )

# 2. Define the offset:
# In Poisson/Negative Binomial regression, we use log(exposure) as an offset
# Exposure = Traffic Volume (ADT) * Length of road
# We add a small constant (or ensure no zeros) to avoid log(0)
penn_model_data <- penn_model_data %>%
  mutate(log_exposure = log(adt12 * length + 1))
```


# Building Model
```{r}
# 3. Fit the Negative Binomial Regression Model
init_model <- glm.nb(tot12 ~ Treat + width + speed + shoulder + offset(log_exposure), 
                data = penn_model_data)

# 4. View the results
summary(init_model)
```

# Model Diagnostics
```{r}
# Check for multicollinearity (ensure your predictors aren't redundant)
car::vif(init_model)

# Check residuals to see if the model fit the data well
plot(init_model)
```
# AIC stepwise selection
```{r}
# Create a model with all possible interactions between Treat and covariates
complex_model <- glm.nb(tot12 ~ Treat * (width + speed + shoulder + curvature) + offset(log_exposure), 
                        data = penn_model_data)

# Simplify using AIC
final_interaction_model <- step(complex_model, direction = "both")
```

To thoroughly investigate whether the efficacy of rumble strips was conditional on roadway geometry, we fitted a complex model including all first-order interactions between the treatment indicator and road characteristics (width, speed, shoulder, curvature). We then employed an automated AIC-based stepwise selection procedure to eliminate non-informative predictors. Throughout this iterative process, all interaction terms were discarded due to a lack of statistical significance and negative impacts on model parsimony. Ultimately, the stepwise procedure again identified the treatment indicator itself as redundant. These findings consistently indicate that, within this dataset, rumble strips were not associated with reduced crash frequency, either as a main effect or through conditional interaction with roadway geometry.

### Selected Prediction Model
```{r}
# The final selected model
final_model <- glm.nb(tot12 ~ width + speed + shoulder + curvature + offset(log_exposure), 
                      data = penn_model_data)

# Summary of the final model
summary(final_model)
```

#DiD model
```{r}
long_data <- penn_model_data %>%
  pivot_longer(
    cols = c(tot04, tot08, tot12, adt04, adt08, adt12),
    names_to = c(".value", "year"),
    names_pattern = "([a-z]+)(\\d+)"
  )
```

```{r}
long_data <- long_data %>%
  mutate(
    post = as.numeric(post),
    Treat = as.factor(Treat),
    year = as.factor(year)
  )
long_data
```


```{r}
did_model <- glm.nb(
  tot ~ Treat * post +
    width + speed + shoulder + curvature +
    offset(log(adt * length)),
  data = long_data
)

summary(did_model)
```

```{r}
did_model2 <- glm.nb(
  tot ~ Treat * year +
    width + speed + shoulder + curvature +
    offset(log(adt * length)),
  data = long_data
)

summary(did_model2)
cat("Converged:", did_model2$converged, "\n")
cat("Fisher scoring iterations:", did_model2$iter, "\n")
cat("Theta:", did_model2$theta, "\n")
cat("SE(theta):", did_model2$SE.theta, "\n")
cat("AIC:", AIC(did_model2), "\n")
```
We estimate a difference-in-differences model using a negative binomial regression to account for overdispersed crash counts. The interaction between treatment status and time captures the causal effect of rumble strip installation. In the baseline DiD model using a single post-treatment indicator, the coefficient on the interaction term (Treat × Post) is -0.168, which corresponds to approximately a 15.5% reduction in crashes (exp(-0.168) ≈ 0.845) for treated road segments relative to untreated segments after installation. However, this effect is not statistically significant (p = 0.198), indicating that the observed reduction could be due to random variation rather than a true treatment effect. Extending the analysis to a more flexible specification with year fixed effects, the estimated coefficient for the treatment effect in 2012 is -0.220, corresponding to approximately a 20% reduction in crashes. However, this effect is not statistically significant (p = 0.141), indicating insufficient evidence to conclude that rumble strips reduce crashes. Importantly, the interaction between treatment status and the pre-treatment year (2008) is also not statistically significant, suggesting no evidence of differential pre-treatment trends between treated and untreated sites and lending support to the parallel trends assumption underlying the DiD approach. Across both models, the estimated treatment effects are consistently negative and of meaningful magnitude, but lack statistical significance. Overall, while the results suggest that rumble strips may reduce crashes, the evidence is not strong enough to draw a definitive causal conclusion.

## Parallel trends
To assess the parallel trends assumption, we examine whether treated and untreated road segments exhibited similar trends in crash counts prior to the installation of rumble strips. In the difference-in-differences model with year fixed effects, the interaction between treatment status and the 2008 indicator is not statistically significant (p = 0.49), indicating no evidence of differential pre-treatment trends between the two groups. This suggests that changes in crashes from 2004 to 2008 were comparable across treated and untreated sites. A visual inspection of average crash counts over time further supports this conclusion, as both groups display broadly similar patterns prior to 2012, with no substantial divergence in trends. Together, these findings support the plausibility of the parallel trends assumption underlying the DiD approach.

# model eval
Most importantly, we examine the validity of the difference-in-differences design through a pre-treatment trend check. The interaction between treatment status and the 2008 indicator is not statistically significant, indicating no evidence of differential pre-treatment trends between treated and untreated road segments. This supports the parallel trends assumption underlying the DiD approach. Additionally, the estimated effects of control variables—such as reduced crashes on wider roads and increased crashes with greater curvature—are consistent with expectations, providing further confidence in the model specification.

```{r}
AIC(did_model2, did_model)
```

```{r}
exp(cbind(
  Estimate = coef(did_model2)[c("Treat1:year08", "Treat1:year12")],
  confint(did_model2)[c("Treat1:year08", "Treat1:year12"), ]
))
```
# Results
```{r}
coef_data <- broom::tidy(did_model2) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    estimate_exp = exp(estimate),
    lower = exp(estimate - 1.96 * std.error),
    upper = exp(estimate + 1.96 * std.error),
    term = reorder(term, estimate_exp)
  )
ggplot(coef_data, aes(x = estimate_exp, y = term)) +
  geom_point() +
  geom_errorbarh(aes(xmin = lower, xmax = upper),
                 height = 0.2) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "red") +
  labs(
    title = "Estimated Effects on Crash Rates (Rate Ratios)",
    x = "Rate Ratio (Multiplicative Effect on Crash Rate)",
    y = ""
  ) +
  theme_minimal()
```

# Sensitivity Analysis
```{r}
# Sensitivity Analysis
library(MASS)

# 1. Define the controls exactly as they appear in the model
controls <- c("width", "speed", "shoulder", "curvature")

# 2. Create a data frame to store results
sensitivity_results <- data.frame(
  dropped_var = c("None", controls),
  est_rate_ratio = NA,
  lower_ci = NA,
  upper_ci = NA
)

# 3. Helper to extract the treatment effect for 2012
get_results <- function(model) {
  term_name <- "Treat1:year12"
  
  # Check that the expected coefficient exists
  if (!(term_name %in% names(coef(model)))) {
    stop(paste("Coefficient", term_name, "not found. Available terms are:",
               paste(names(coef(model)), collapse = ", ")))
  }
  
  rate_ratio <- exp(coef(model)[term_name])
  conf_int <- exp(confint(model)[term_name, ])
  
  return(c(rate_ratio, conf_int))
}

# 4. Store the original model results
sensitivity_results[1, 2:4] <- get_results(did_model2)

# 5. Loop to drop one control variable at a time
for (i in seq_along(controls)) {
  vars_to_keep <- controls[controls != controls[i]]
  
  formula_str <- as.formula(
    paste(
      "tot ~ Treat * year +",
      paste(vars_to_keep, collapse = " + "),
      "+ offset(log(adt * length))"
    )
  )
  
  # Re-fit the model
  temp_model <- glm.nb(formula_str, data = long_data)
  
  # Store results
  sensitivity_results[i + 1, 2:4] <- get_results(temp_model)
}

# 6. View the final table
print(sensitivity_results)
```
The estimated treatment effect is extremely stable across model specifications.

A sensitivity analysis was conducted by re-estimating the model while sequentially excluding each control variable. The estimated treatment effect remains highly stable across all specifications, with rate ratios consistently near 0.80 and nearly identical confidence intervals. This indicates that the estimated effect of rumble strip installation is not driven by any single control variable and provides additional support for the robustness of the model specification.
