---
title: "EDA"
format: PDF
editor: visual
---

```{r}
library(tidyverse)
```

```{r}
penn_crash <- read.table("PennCrash.txt", header = TRUE, quote = "\"", sep = "")

penn_crash
```
```{r}
sum(penn_model_data$Treat == 1)
```
only 331 treated observations compared to 1,655 treated.

# Crash Rate Trend Plot
```{r}
# 1. Calculate yearly rates
penn_crash <- penn_crash %>%
  mutate(rate04 = tot04 / (adt04 * length),
         rate08 = tot08 / (adt08 * length),
         rate12 = tot12 / (adt12 * length))

# 2. Summarize the mean rates by year and treatment
yearly_rate_summary <- penn_crash %>%
  group_by(Treat) %>%
  summarise(mean_rate04 = mean(rate04, na.rm = TRUE),
            mean_rate08 = mean(rate08, na.rm = TRUE),
            mean_rate12 = mean(rate12, na.rm = TRUE))

print(yearly_rate_summary)

# 3. Optional: Plot the trend of these rates
rate_long <- yearly_rate_summary %>%
  pivot_longer(cols = starts_with("mean_rate"), names_to = "year", values_to = "rate") %>%
  mutate(year = recode(year, "mean_rate04" = 2004, "mean_rate08" = 2008, "mean_rate12" = 2012))

ggplot(rate_long, aes(x = factor(year), y = rate, color = factor(Treat), group = Treat)) +
  geom_line(size = 1) +
  geom_point(size = 3) +
  labs(title = "Average Yearly Crash Rate",
       x = "Year", y = "Crashes per Traffic-Mile",
       color = "Treatment") +
  theme_minimal()
```
Figure X displays average crash rates (crashes per traffic-mile) over time for treated and untreated road segments. Prior to treatment (2004–2008), both groups exhibit similar upward trends, providing visual support for the parallel trends assumption. Following the installation of rumble strips, treated sites experience a substantially larger decline in crash rates compared to untreated sites. This pattern is consistent with a potential beneficial effect of rumble strips. However, consistent with the regression results, this difference is not statistically significant and should be interpreted with caution.

# Confounder Analysis
```{r}
# Check for balance across key categorical variables
# We use 'pivot_longer' to create a faceted plot of all categorical variables
categorical_vars <- c("width", "speed", "shoulder")

penn_crash %>%
  select(Treat, all_of(categorical_vars)) %>%
  pivot_longer(cols = -Treat, names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = value, fill = factor(Treat))) +
  geom_bar(position = "fill") + # 'fill' makes them percentage bars for easier comparison
  facet_wrap(~variable, scales = "free") +
  labs(title = "Proportional Balance of Road Characteristics by Treatment",
       y = "Proportion of Sites", x = "Category", fill = "Treatment") +
  theme_minimal()
```
The proportional bar charts reveal that rumble strip installations were not assigned randomly with respect to site characteristics. Specifically, treatment sites appear disproportionately represented in certain categories—such as higher-speed zones and specific pavement widths—compared to control sites. Because these characteristics (speed, width, shoulder) are independently associated with crash risk, the lack of balance indicates that simple comparisons between the treatment and control groups may be biased by these pre-existing structural differences. Consequently, these variables must be included as covariates in the regression model to isolate the effect of the rumble strips from the influence of road geometry.


#_____________________________________________________________________________________________________________________

### Distribution of Crashes
```{r}
ggplot(penn_model_data, aes(x = tot12)) +
  geom_histogram(bins = 30) +
  labs(title = "Distribution of Crash Counts (2012)",
       x = "Number of Crashes",
       y = "Frequency") +
  theme_minimal()
```
Right skewed, many zeros

### Mean vs Variance
```{r}
mean(penn_model_data$tot12)
var(penn_model_data$tot12)
```
variance >> mean
The variance of crash counts exceeds the mean, indicating overdispersion and motivating the use of a negative binomial model.

### Traffic Exposure Trends
```{r}
binned_data <- penn_model_data %>%
  mutate(exposure = adt12 * length / 1000,
         exposure_bin = cut(exposure, breaks = 10)) %>%
  group_by(exposure_bin) %>%
  summarise(
    mean_crashes = mean(tot12, na.rm = TRUE),
    midpoint = mean(exposure, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(binned_data, aes(x = midpoint, y = mean_crashes)) +
  geom_point() +
  geom_line() +
  labs(
    title = "Average Crash Counts Increase with Traffic Exposure",
    x = "Traffic Exposure (thousands)",
    y = "Average Crashes"
  ) +
  theme_minimal()
```
Average crash counts increase with traffic exposure, though the relationship is not perfectly linear. Higher exposure bins are associated with higher mean crash counts, supporting the inclusion of an exposure offset.




